---
title: "Diagnostics Report: Bayesian Hierarchical Abudance Model"
author: "Valentina La Morgia"
date: last-modified
format: 
  html:
    toc: true
    code-fold: true
    self-contained: true
knitr:
  opts_chunk: 
    fig.width: 10   # Larghezza in pollici
    fig.height: 7   # Altezza in pollici
    fig.align: "center"
    dpi: 300        # Alta risoluzione
---


```{r}
#| label: packages
#| message: false
#| warning: false
##---- Load required packages ----
{
  library(cmdstanr)
  library(posterior)
  library(bayesplot)
  library(corrplot)
  library(vegan)
  library(sf)
  library(patchwork)
  library(tidybayes)
  library(bayestestR)
  library(overlapping)
}
```

```{r}
#| label: reconstruct-model
#| message: false
#| warning: false
rm(list = ls())
source("scripts/03_abundance_hierachical_STAN_data_prep.R")
fit <- readRDS("outputs/rds/fit_mod_inform.rds")
```


## 1. Convergence
In this section, we verify that the model has correctly explored the parameter space.   
Rhat must always be less than 1.01 and ESS > 400.
In general, there are no divergences, we have no max treedepth, and the ebfmi value is good (it must not be < 0.2/0.3).

```{r}
#| label: summary-diagnostics

fit$diagnostic_summary()
# fit$cmdstan_diagnose()

####---- Convergence ----
# Main diagnostics for the main parameters
summary_diag <- fit$summary(c(
  "psi_area",   # factor for areas
  "alpha_p",    # detection
  "beta",       # recruitment (births and immigration)
  "phi",        # survival, derived from global parameter mu (mortality)
  "N"           # abundance
))
print(summary_diag %>% select(variable, rhat, ess_bulk, ess_tail))
print(paste0("Rhat is always smaller than: ", round(max(summary_diag$rhat),3)))
print(paste0("ESS_bulk is always larger than: ", round(min(summary_diag$ess_bulk),0)))
print(paste0("ESS_tail is always larger than: ", round(min(summary_diag$ess_tail),0)))

# Diagnostic plot: Rhat vs ESS
ggplot(summary_diag, aes(x = rhat, y = ess_bulk)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  geom_vline(xintercept = 1.01, linetype = "dashed", color = "red") +
  labs(title = "Diagnostica MCMC: Rhat vs ESS",
       subtitle = "La linea rossa indica la soglia critica di Rhat (1.05)",
       x = "Rhat (Convergenza)",
       y = "Bulk ESS (Efficienza)") +
  theme_minimal()
```

The validity of the MCMC sampling was confirmed by analyzing the values of $\hat{R}$ (Gelman-Rubin statistic), all of which were below the critical threshold of 1.01, indicating correct convergence of the chains. Furthermore, the Effective Sample Size (ESS) for the parameters of interest ($N$, $\phi$, $\beta$) exceeded the minimum recommended threshold, ensuring sufficient accuracy for the estimation of credibility intervals.

The following graphs should show ‘intertwined caterpillars’.

```{r}
#| label: mcmc-trace-N
#| fig-cap: "MCMC Diagnostics: Traceplots - Abundance"
# Trace plots to visualise convergence

# for N, we select an area with low density (S1) and one with high density (R3), and for both we plot the first session and one in the middle
mcmc_trace(fit$draws(c("N[1,1]", "N[1,20]", "N[6,1]", "N[6,20]")))
```

For the abundance parameter, we only plotted an area with low density (S1) and one with high density (R3), and for both we selected the first session and one in the middle/towards the end. Trace plots for the specific abundance parameters (e.g., N[1,1], N[6,20]) demonstrate good convergence and mixing. The y-axis scales reflect the varying densities between areas; for instance, the estimates for area 6 reach up to approximately 80 individuals, while area 1 estimates are significantly lower, staying below 12. Recall that for N, the caterpillars may appear slightly “wider” (more uncertainty) for areas with few captures (such as area S1 in this case). This is normal, as long as the chains overlap well.

```{r}
#| label: mcmc-trace-alpha_p
#| fig-cap: "MCMC Diagnostics: Traceplots - Detectability"
# Trace plots to visualise convergence
mcmc_trace(fit$draws(c("alpha_p[1,1]", "alpha_p[1,11]", "alpha_p[1,22]",
                     "alpha_p[6,1]", "alpha_p[6,11]", "alpha_p[6,22]")))
mcmc_trace(fit$draws(c("alpha_p[1,1]", "alpha_p[1,2]", "alpha_p[1,3]")))
```

For alpha_p (detection), it is useful to check convergence at the three “extremes” of monitoring to ensure that the model is stable throughout the time frame. Again, we used areas S1 (1) and R3 (6) (different abundance), selecting an initial session (1), an intermediate session (11), and a final session (22).    
Another critical point of the model is how it estimates detectability in sites with many zeros. Checking the trace plots of these intercepts is essential to see if the “borrowing strength” (information sharing) is working well or if the chains are wandering aimlessly.     
The graphs for areas 1 and 6 (e.g., alpha_p[1,1], alpha_p[6,22]) show perfectly overlapping chains. This indicates that the model has reached a stable estimate of the probability of capture for different sessions and areas.
With regard to the management of scarce data, the plots alpha_p[1,1], alpha_p[1,2], and alpha_p[1,3] are particularly interesting. These correspond to area S1 in the first sessions of 2022, where the tables indicate zero catches. Even in the total absence of direct observations in those years, the chains converge. This happens because the hierarchical model “borrows” information from other sites and years, stabilizing the estimate through global distribution.
Regarding detectability variability, note how the mean values of the intercepts (on the y-axis) vary between sessions. For example, alpha_p[6,1] has a higher mean than alpha_p[6,11], correctly reflecting how the probability of capture changes over time depending on sampling conditions.

```{r}
#| label: mcmc-trace-beta
#| fig-cap: "MCMC Diagnostics: Traceplots - Entry probability"
# Trace plots to visualise convergence

mcmc_trace(fit$draws(c(
  "beta[6,12]", # Area 6, ripresa Maggio 2023 (Post-Inverno)
  "beta[6,19]", # Area 6, ripresa Giugno 2024 (Post-Inverno)
  "beta[6,5]",  # Area 6, Luglio 2022 (Dati abbondanti)
  "beta[6,15]",  # Area 6, Luglio 2022 (Dati abbondanti)
  "beta[1,2]",  # Area 1, Maggio 2022 (No avvistamenti)
  "beta[1,5]",  # Area 1, Luglio 2022 (Dati scarsi)
  "beta[1,12]"  # Area 1, Maggio 2023 (Incertezza massima)
)))
mcmc_trace(fit$draws(c("phi")))
```

Given the fragmented sampling schedule (with long winter breaks), the $\beta$ parameters (entry probability) are particularly interesting because they have to handle both “constant” recruitment during the summer and the massive entry of new individuals after winter. 
We have therefore chosen the $\beta$ values that correspond to the first session after the winter break. Here, the model must estimate who “entered” during the 7 months of absence. $\beta[s, 12]$: Corresponds to May 15, 2023 (first entry after winter 2022). $\beta[s, 19]$: Corresponds to June 15, 2024 (first entry after winter 2023). The chains here must be well mixed, but we can expect higher values than the others.    
Results for beta[6,12] and beta[6,19]: In these post-winter recovery sessions, the chains are well mixed and stationary. In particular, beta[6,19] shows oscillations covering almost the entire interval $[0, 1]$, indicating that the model recognizes a significant influx of new individuals, but with inherent uncertainty due to the long sampling break that preceded it.     
We then chose a mid-summer session (July/August) in an area with a lot of data (e.g., R3). $\beta[6, 5]$ (Area 6, July 2022) or $\beta[6, 15]$ (Area 6, August 2023). Here we expect very thin and precise “caterpillars” because with a lot of capture data, the model should be very confident about when new individuals enter.     
Results for $\beta[6, 5]$ and $\beta[6, 15]$: These plots correspond to summer sessions (July 2022 and August 2023) in the R3 area. The chains are much “narrower” and concentrated towards low values (e.g., below 0.20 for beta[6,5]). This indicates that, with close sampling and more data, the model is very accurate in identifying that summer recruitment is constant but limited compared to spring peaks.      
Finally, it is essential to show convergence in sites where we had many zeros. $\beta[1, 2]$ (Area 1, start of study) is useful to show how the model handles recruitment when it sees no one. The chains are “wider” (greater uncertainty), but they still overlap. This shows that the model does not “go crazy” in the absence of data, but correctly relies on the hierarchical structure.     
Results for $\beta[1,2]$, $\beta[1,5]$, and \$beta[1,12]$: These graphs refer to area S1, which had no catches in several occasions. These trace plots are very “thick” and cover almost the entire space from 0 to 1. This is correct behavior: in the absence of local data, the model expresses maximum uncertainty, while remaining stable thanks to information “borrowed” from the global mortality rate ($\mu$).

```{r}
#| label: mcmc-trace-phi
#| fig-cap: "MCMC Diagnostics: Traceplots - Survival"
# Trace plots to visualise convergence

mcmc_trace(fit$draws(c("phi")))
```

Finally, the trace plots for $\phi$ (survival, derived from the global $\mu$ parameter, mortality) show generally high values (often between 0.90 and 0.96), indicating high apparent survival between sessions. Note how the intervals phi[11] and phi[18] show lower values and greater variance (ranging between 0.2 and 0.6). Since it is derived from a global parameter ($\mu$), the trace plots of phi should be, as in this case, very clean and stable. If they fluctuate greatly, on the other hand, it could mean that survival data is scarce in all areas.


## 2. Prior-Posterior overlap

To verify the overlap between priors and posteriors, we considered the global parameter $\mu$ (active and hibernation periods), and for the probability of inclusion per area, detection, and betas, we compared a data-rich area (R3) and a data-poor area (S1). 

```{r}
#| label: Prior-posterior-overlap-mu_active
#| fig-cap: "Prior-Posterior overlap: global mortality - active period"

# pior samples (mu ~ half-normal(0,1))
set.seed(123)
n_samples <- 12000 # Total number of samples (3000 iterations * 4 chains)
prior_mu <- abs(rnorm(n_samples, 0, 1))
# extract the posterior
post_mu <- as.vector(fit$draws("mu_active"))

df_mu <- data.frame(
  Value = c(prior_mu, post_mu),
  Type = rep(c("Prior", "Posterior"), each = n_samples)
)
ggplot(df_mu, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.5) +
  labs(title = "Prior-Posterior Overlap: Global Mortality (mu_active)",
       x = "Mortality", y = "Density") +
  theme_minimal()
```

```{r}
#| label: Prior-posterior-overlap-mu_hibernation
#| fig-cap: "Prior-Posterior overlap: global mortality - hibernation period"

# pior samples (mu ~ half-normal(0,1))
set.seed(123)
n_samples <- 12000 # Total number of samples (3000 iterations * 4 chains)
prior_mu <- abs(rnorm(n_samples, 0, 1))
# extract the posterior
post_mu <- as.vector(fit$draws("mu_hibernation"))

df_mu <- data.frame(
  Value = c(prior_mu, post_mu),
  Type = rep(c("Prior", "Posterior"), each = n_samples)
)
ggplot(df_mu, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.5) +
  labs(title = "Prior-Posterior Overlap: Global Mortality (mu_hibernation)",
       x = "Mortality", y = "Density") +
  theme_minimal()
```

The comparison between the prior and posterior distributions for the global mortality rate (μ) shows minimal overlap: the reduction in variance in the posterior distribution compared to the prior distribution confirms that the experimental data were highly informative, allowing an accurate estimate of the parameter through the hierarchical structure of the model.


```{r}
#| label: Prior-posterior-overlap-psi_areas
#| fig-cap: "Prior-Posterior overlap: inclusion probability (psi)"

prior_psi <- rbeta(n_samples, 1, 1)
post_psi_ricca <- as.vector(fit$draws("psi_area[6]"))
post_psi_povera <- as.vector(fit$draws("psi_area[1]"))

df_psi <- data.frame(
  Value = c(prior_psi, post_psi_ricca, post_psi_povera),
  Type = rep(c("Prior", "Post: R3 (data-rich)", "Post: S1 (data-poor)"), each = n_samples)
)
ggplot(df_psi, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.4) +
  labs(title = "Prior-Posterior Overlap: Inclusion probability (psi)",
      #  subtitle = "Evidenzia come i dati locali riducono l'incertezza rispetto alla prior",
       x = "logit(psi)", y = "Density") +
  theme_minimal()

```

The inclusion probability per area ($\psi_{area}$) determines how many of the increased individuals ($M$) are real. Having used a $\text{Beta}(1, 1)$ (a flat line between 0 and 1), any “hump” in the posterior indicates that the model has learned something about the population size.     
This graph confirms that the choice of data augmentation value (M) was correct: since none of the posterior distributions are "squashed" against the upper limit of 1, we have proof that the value of M did not artificially limit the estimate of the actual abundance. Furthermore, it graphically demonstrates borrowing strength: even in the data-poor site (S1), the model manages to define a plausible distribution thanks to the influence of the global hierarchical structure.

```{r}
#| label: Prior-posterior-overlap-alpha_p
#| fig-cap: "Prior-Posterior overlap: global mortality"

prior_alpha <- rnorm(n_samples, 0, 2)
post_alpha_ricca <- as.vector(fit$draws("alpha_p[6,11]"))
post_alpha_povera <- as.vector(fit$draws("alpha_p[1,11]"))

df_alpha <- data.frame(
  Value = c(prior_alpha, post_alpha_ricca, post_alpha_povera),
  Type = rep(c("Prior", "Post: R3 (data-rich)", "Post: S1 (data-poor)"), each = n_samples)
)
ggplot(df_alpha, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.4) +
  labs(title = "Prior-Posterior Overlap: Detectability (alpha_p)",
      #  subtitle = "Evidenzia come i dati locali riducono l'incertezza rispetto alla prior",
       x = "logit(p)", y = "Density") +
  theme_minimal()
```

The detection intercept ($\alpha_p$) is the most sensitive parameter in capture-recapture models. In the data-rich area, we expect and indeed observe minimal overlap (the data dominates).

```{r}
#| label: Prior-posterior-overlap-beta
#| fig-cap: "Prior-Posterior overlap: recruitment (beta)"

prior_beta <- rbeta(n_samples, 1, 1)
post_beta_ricca <- as.vector(fit$draws("beta[6,11]"))
post_beta_povera <- as.vector(fit$draws("beta[1,11]"))

df_beta <- data.frame(
  Value = c(prior_beta, post_beta_ricca, post_beta_povera),
  Type = rep(c("Prior", "Post: R3 (data-rich)", "Post: S1 (data-poor)"), each = n_samples)
)
ggplot(df_beta, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.4) +
  labs(title = "Prior-Posterior Overlap: Recruitment (beta)",
      #  subtitle = "Evidenzia come i dati locali riducono l'incertezza rispetto alla prior",
       x = "beta", y = "Density") +
  theme_minimal()
```

The entry probability ($\beta$) represents recruitment, which is why we have again chosen the $\beta$ values from the post-winter sessions (e.g., session 11 or 18). Here too, starting from a $\text{Beta}(1, 1)$, we expect the model to “shift” the probability towards values that better explain the appearance of new individuals.     
This graph validates the hierarchical structure of the model. It confirms that the global mortality rate is not influenced by your initial choices (the priors), but is driven almost exclusively by the biological evidence that emerged from the six areas of study.

### 2.1. Prior-Posterior Overlap for N

For the parameter $N$ (Abundance), the Prior-Posterior Overlap (PPO) check has a different meaning and, in some ways, is less “canonical” than structural parameters such as $\mu$ or $\alpha_p$. In our model, $N$ is not a parameter to which we directly assign a prior (as we do with $\mu \sim \text{Normal}(0,1)$). $N$ is calculated in the generated quantities block as the sum of the individual probabilities of being alive. The “prior” of $N$ is implicit: the prior distribution of $N$ depends on the combination of the priors of $\psi$ (inclusion), $\beta$ (entry), and $\phi$ (survival). This “prior” of $N$ is bounded above by the value $M$ (data augmentation) chosen in the R code ($M = n \times 3$). However, it may make sense to perform a PPO check to demonstrate that the value $M$ chosen for augmentation is sufficient. If the posterior of $N$ is very close to $M$, it means that the prior is cutting off reality and it is necessary to increase $M$.


```{r}
#| label: Prior-posterior-overlap-N
#| fig-cap: "Prior-Posterior overlap: Abundance (N)"

post_N <- as.vector(fit$draws("N[6,11]"))
# Definition of the prior (simplified)
# if psi and beta are uniform, N follows a distribution related to M
prior_N <- runif(length(post_N), 0,
  stan_data$M_total /   stan_data$N_areas) 

df_N <- data.frame(
  Value = c(prior_N, post_N),
  Type = rep(c("Prior (Implicita/M)", "Posterior"), each = length(post_N))
)
ggplot(df_N, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.5) +
  labs(title = "Overlap for Abundance (N)",
      #  subtitle = "Dimostra quanto i dati hanno ristretto la stima rispetto al limite M",
       x = "Number of individuals (N)", y = "Density") +
  theme_minimal()
```

The overlap check for the abundance parameter (N) confirms the robustness of the data augmentation strategy. The posterior distribution is concentrated in a much narrower range of values than the implicit prior, with a tail that ends well before the maximum limit M. This shows that the monitoring provided sufficiently informative data to allow the model to converge on a population estimate that is not influenced by the numerical constraints of the augmented dataset.

### 2.2. Prior-Posterior Overlap for Survival (phi)

Survival in our model is a transformed parameter that depends on two elements: the overall mortality rate $\mu$ and the time interval $\Delta$ (delta). Calculating the PPO for $\phi$ is still useful: while $\mu$ is an instantaneous rate that is difficult to visualize, $\phi$ is an intuitive probability (0-1). Showing the overlap for $\phi$ allows us to see how the “prior” on $\mu$ translates into a “prior” on the actual survival of the animals. Furthermore, since we have very different intervals $\Delta$ (15 days vs. 7 months), the “prior” of $\phi$ changes for each session. Showing the PPO for a short and a long interval should demonstrate that the model understands that the probability of surviving one winter is different from that of surviving two weeks.

```{r}
#| label: Prior-posterior-overlap-survival
#| fig-cap: "Prior-Posterior overlap: Survival (phi)"

# temporal intervals
# delta[1] ~15 gg, delta[11] winter (~210 gg)
d_breve <- 15 
d_lungo <- 210

# generate phi prior transforming the mu prior ~ HalfNormal(0,1)
prior_mu <- abs(rnorm(n_samples, 0, 1))
prior_phi_breve <- exp(-prior_mu * (d_breve/30.44)) # normalised on a monthly basis (cf. intervals)
prior_phi_lungo <- exp(-prior_mu * (d_lungo/30.44))
post_phi_breve <- as.vector(fit$draws("phi[1]"))
post_phi_lungo <- as.vector(fit$draws("phi[18]"))

# Funzione per calcolare il picco (moda) della densità
get_mode <- function(x) {
  dens <- density(x)
  return(dens$x[which.max(dens$y)])
}
# Calcola i valori dei picchi
picco_breve <- get_mode(post_phi_breve)
picco_lungo <- get_mode(post_phi_lungo)
# Stampa i risultati
cat("Picco sopravvivenza breve (~15gg):", round(picco_breve, 3), "\n")
cat("Picco sopravvivenza invernale (~7 mesi):", round(picco_lungo, 3), "\n")

df_phi_short <- data.frame(
  Value = c(prior_phi_breve, post_phi_breve),
  Type = rep(c("Prior", "Posterior"), each = length(post_N))
)
ggplot(df_phi_short, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.5) +
  labs(#title = "Overlap for Survival (phi) - short interval",
       subtitle = "Short summer interval",
       x = "Survival", y = "Density") +
  theme_minimal() -> df_phi_short

df_phi_long <- data.frame(
  Value = c(prior_phi_lungo, post_phi_lungo),
  Type = rep(c("Prior", "Posterior"), each = length(post_N))
)

ggplot(df_phi_long, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.5) +
  labs(#title = "Overlap for Survival (phi) - long interval",
       subtitle = "Long winter interval)",
       x = "Survival", y = "Density") +
  theme_minimal() -> df_phi_long

df_phi_short + df_phi_long +
  plot_layout(guides = "collect") + 
  plot_annotation(title = "Prior-Posterior Overlap for Survival (phi)")
```

```{r}
# 1. Definizione degli indici stagionali (basata sui tuoi intervalli)
idx_summer <- c(1:10, 12:17, 19:22)
idx_winter <- c(11, 18)

# 2. Generazione delle Prior aggregate
# Usiamo la distribuzione di mu per simulare cosa si aspetterebbe il modello "al buio"
mu_prior_samples <- abs(rnorm(n_samples, 0, 1))

# Funzione per scalare la prior di mu sugli intervalli reali (assumendo 15gg estate, 210gg inverno)
prior_phi_summer_all <- as.vector(sapply(c(15), function(d) exp(-mu_prior_samples * (d/30.44))))
prior_phi_winter_all <- as.vector(sapply(c(210), function(d) exp(-mu_prior_samples * (d/30.44))))

# 3. Estrazione delle Posterior aggregate dal fit
post_phi_summer_all <- as.vector(fit$draws(paste0("phi[", idx_summer, "]")))
post_phi_winter_all <- as.vector(fit$draws(paste0("phi[", idx_winter, "]")))

# 4. Funzione per il calcolo della Moda (Picco)
get_mode <- function(x) {
  dens <- density(x)
  return(dens$x[which.max(dens$y)])
}

picco_summer <- get_mode(post_phi_summer_all)
picco_winter <- get_mode(post_phi_winter_all)

# 5. Calcolo dell'indice di Overlap numerico (OV)
# Un valore basso (es. 0.20) indica che i dati hanno cambiato radicalmente la prior
ov_summer <- overlapping::overlap(list(prior_phi_summer_all, post_phi_summer_all))$OV
ov_winter <- overlapping::overlap(list(prior_phi_winter_all, post_phi_winter_all))$OV

# --- STAMPA RISULTATI PER IL PAPER ---
cat("--- Risultati Aggregati Sopravvivenza (phi) ---\n")
cat("ESTATE - Picco:", round(picco_summer, 3), "| Overlap Prior-Post:", round(ov_summer, 3), "\n")
cat("INVERNO - Picco:", round(picco_winter, 3), "| Overlap Prior-Post:", round(ov_winter, 3), "\n")

# 6. Creazione dei Plot Aggregati con Patchwork
df_summer <- data.frame(
  Value = c(
    sample(prior_phi_summer_all, length(post_phi_summer_all), replace = TRUE), 
    post_phi_summer_all
  ),
  Type = rep(c("Prior", "Posterior"), each = length(post_phi_summer_all))
)

df_winter <- data.frame(
  Value = c(
    sample(prior_phi_winter_all, length(post_phi_winter_all), replace = TRUE), 
    post_phi_winter_all
  ),
  Type = rep(c("Prior", "Posterior"), each = length(post_phi_winter_all))
)

p1 <- ggplot(df_summer, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = picco_summer, linetype = "dashed") +
  labs(subtitle = paste0("Summer Aggregate (OV: ", round(ov_summer, 2), ")"),
       x = "Survival (phi)", y = "Density") +
  theme_minimal()

p2 <- ggplot(df_winter, aes(x = Value, fill = Type)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = picco_winter, linetype = "dashed") +
  labs(subtitle = paste0("Winter Aggregate (OV: ", round(ov_winter, 2), ")"),
       x = "Survival (phi)", y = "Density") +
  theme_minimal()

p1 + p2 + 
  plot_layout(guides = "collect") + 
  plot_annotation(title = "Aggregate Prior-Posterior Overlap for Seasonal Survival")

ggsave(
    filename = "figs/Prior_Post_overlap_phi.tiff", # o .png, .pdf, .eps
    plot = last_plot(),                          # esporta l'ultimo grafico visualizzato
    device = "tiff",                             # formato file
    compression = "lzw",
    width = 180,                                 # larghezza (es. 180mm è lo standard "full page width")
    height = 120,                                # altezza in mm
    units = "mm",                                # unità di misura
    dpi = 300,                                   # risoluzione richiesta
    bg = "white"                                 # sfondo bianco (evita trasparenze indesiderate)
  )

```

The comparison of overlap checks for apparent survival ($\Phi$) between short and long intervals highlights the model's ability to adapt to the temporal structure of the data. While survival is estimated with extreme precision close to unity in 15-day intervals, the model correctly captures the decrease in survival probability in inter-annual intervals. The clear distinction between the posterior distributions in the two scenarios confirms that the monthly mortality rate $\mu$ has been estimated robustly, allowing for consistent demographic projections on both a biweekly and seasonal scale.

However, for the long interval (winter) survival, we notice that the peak of the posterior is around 0.75 - 0.80, and the distribution is much "fatter" and lower than the short interval.
In the Bayesian model, the hierarchical structure (sharing information across sites) is "pulling" the winter survival upward (toward 0.80) to be more consistent with the rest of the year. If the model insists that survival is relatively high (0.80) even in winter, but we still don't see any animals in the spring, the model has only one mathematical way to explain the missing data: the animals must have never entered the population or the super-population itself must be smaller. The posterior check is valid, but it reveals that the global "borrowing of strength" might be slightly over-estimating winter survival compared to what the raw data suggests.    
So, this confirms the "conservative" nature of our Bayesian estimates. While the model finds a high precision for short-term survival, it acknowledges greater uncertainty during the winter, where the hierarchical structure provides a more stable (though perhaps more optimistic) estimate of survival than frequentist methods.

## 3. Correlation between parameters

In a capture-recapture model, parameters are almost never independent. Seeing how they relate to each other helps to understand whether the model is well specified or whether there are identifiability issues.

The correlation between abundance and detectability is the most important one. It is often negative: if the model estimates low detectability, it must compensate by estimating more individuals to explain the observed data (and vice versa).    
If the correlation is extreme (e.g. >0.9), it means that the model struggles to distinguish whether there are a few highly visible animals or many nearly invisible animals. If the correlation is moderate, the model is robust.

Between Survival ($\Phi$) and Recruitment ($\beta$), we expect a trade-off.
This correlation helps to understand population dynamics. If an area has high survival but low recruitment, the population is stable but "old". Seeing whether these parameters are correlated in the estimates tells us whether the model correctly separates those who remain (survivors) from those who enter (newcomers).

The correlation between global parameters (e.g. $\mu$) and local parameters allows to see how much the estimates for individual sites depend on the global parameter. It is visual proof of the "borrowing strength" of the hierarchical model.

```{r}
#| label: correlation-abundance-detection
#| fig-cap: "Correlation of abundance and detectability"
params_check <- c("mu_active", "mu_hibernation",
  "psi_area[6]",
  "N[6,11]",
  "alpha_p[6,11]",
  "beta[6,12]")
mcmc_pairs(fit$draws(params_check), off_diag_fun = "hex")
```

There is a clear negative correlation between the two mortality rates. This is a common "trade-off" in capture-recapture: if the model increases mortality in one season, it must adjust the other to fit the total number of surviving individuals.

The analysis of correlations between parameters (Pairs Plot) confirms the identifiability of the model. With the exception of the intrinsic and expected correlation between the probability of inclusion and abundance (N), the other key parameters, in particular the mortality rates and detectability ($\alpha_{p}$), show substantial independence. The absence of 'banana' or funnel shapes in the joint distributions ensures that the MCMC sampler has correctly explored the parameter space without incurring numerical trade-offs that could have affected the accuracy of the demographic estimates.

## 4. Shrinkage and overconfidence

Shrinkage is the phenomenon whereby estimates for individual sites are "pulled" towards the global mean.
It demonstrates the usefulness of the hierarchical model. At sites with very little data (such as S1), the abundance estimate does not "explode" towards absurd values, but is stabilised by the behaviour of the other sites.     
With high shrinkage, in data-poor sites, the posterior will be very close to the global mean. This is good: it means that the model is compensating for the lack of local data.     
With low shrinkage, in data-rich sites (such as R3), the estimate remains closer to the raw data. This indicates that the local data is strong enough to 'win' against the pressure of the global mean.

Overconfidence occurs when the model produces credibility intervals that are too narrow and do not reflect the actual variability of the data, often due to an overly rigid hierarchical structure or overly informative priors.     
Chechking for overconfidence protects us from accusations of having 'forced' the results. If the model claims pinpoint accuracy at a site where you have only captured two animals, you are in a situation of overconfidence.

In our case, we are interested in how N is affected by the hierarchical model. However, this effect is due to the combined effect on alpha_p and psi. So, we eveluate the shrinkage on these two parameters.

Detectability is the parameter where shrinkage is most evident. Without the hierarchical model, a site with very few captures would give an estimate of detectability that is extreme or very unstable.    
We compare the average estimate of alpha_p for each area relative to the global mean (the hyperparameter of the model). Areas with less data (such as S1) should have an alpha_p that 'leans' towards the global mean more than they would if estimated on their own, because we are assuming that the species has similar detectability at all sites, unless the local data are strong enough to prove otherwise.

```{r}
#| label: shrinkage-detectability
#| fig-cap: "Shrinkage: Detectability (p)"

# avarage p across all areas
mean_alpha_all <- fit$summary("alpha_p") %>% 
  pull(mean) %>% 
  mean()
# avarage psi across all areas
mean_psi_all <- fit$summary("psi_area") %>% 
  pull(mean) %>% 
  mean()

alpha_data <- fit$summary("alpha_p") %>%
  mutate(Area = rep(1:6, each = 23)) %>% # 23 is the number of sessions
  group_by(Area) %>%
  summarise(Media_Area = mean(mean), 
            Incertezza_inf = mean(q5), 
            Incertezza_sup = mean(q95))
ggplot(alpha_data, aes(x = factor(Area), y = Media_Area)) +
  geom_hline(yintercept = mean_alpha_all, linetype = "dashed", color = "darkgrey") +
  geom_errorbar(aes(ymin = Incertezza_inf, ymax = Incertezza_sup), width = 0.2, color = "royalblue") +
  geom_point(size = 3, color = "royalblue") +
  labs(title = "Shrinkage: Detectability (p)",
       subtitle = "The dashed line is the mean detectability across all sites",
       x = "Study area", y = "Estimated value (logit scale)") +
  theme_minimal()
```

The graph is visual proof that the hierarchical model works. The dots (the area averages) all gravitate around the dotted line (the average of all sites). Even Area 1 and Area 6, despite their differences in density, have comparable detectability estimates.

On the other hand, psi determines the size of the actual 'super-population'. It is the parameter that most directly influences N.
Shrinkage on the probability of inclusion acts as a density regulariser. It prevents the estimated population in a "poor" area from being too different from the average of other areas, unless there is a solid statistical reason. It shows that the model is not just counting animals, but estimating a probability distribution consistent across all sites.


```{r}
#| label: shrinkage-inclusion
#| fig-cap: "Shrinkage: Inclusion probability (psi)"

mean_psi_all <- fit$summary("psi_area") %>% 
  pull(mean) %>% 
  mean()
psi_data <- fit$summary("psi_area") %>%
  mutate(Area = factor(1:6)) %>% 
  select(Area, mean, q5, q95)
ggplot(psi_data, aes(x = Area, y = mean)) +
  geom_hline(yintercept = mean_psi_all, linetype = "dashed", color = "darkgrey", size = 0.8) +
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0.2, color = "seagreen", size = 1) +
  geom_point(size = 4, color = "seagreen") +
  scale_y_continuous(limits = c(0, 1)) + # psi è una probabilità 0-1
  labs(title = "Shrinkage: Inclusion probability (psi)",
       subtitle = "The dashed line is the mean inclusione probability across all sites",
       x = "Study area", y = "Probability (psi)") +
  theme_minimal()
```

In this graph, the difference in uncertainty between the areas is also clearly visible, confirming that the model does not suffer from overconfidence.    
The error bars for Areas 1 and 2 are significantly wider than those for Area 3 or Area 6. This is because psi is directly proportional to the number of unique individuals captured: the fewer individuals you have observed in an area, the more uncertain the model is about the actual size of the population at that site.    
Concerning shrinkage behaviour, the points in Areas 1 and 2 are 'pulled' upwards (towards the grey dotted line) compared to what a simple raw count would do. The model is saying, 'I saw few animals in these areas, but since the density is higher in the other sites, it is likely that there are more individuals here than I was able to capture'.
Area 6 (Data-rich) is the only area that stands out above the global average with reduced uncertainty. Here, the data is so strong that it 'overcomes' the force of gravity of shrinkage.

Finally, we check for the overall effect on N, by comparing the model estimate with the observed number individuals for a sample session (session 9, 5 Sept. 2022).

```{r}
#| label: shrinkage-N
#| fig-cap: "Shrinkage: Abundance (N)"

k = 9
n_observed <- sapply(1:6, function(a) {
  # individuals captured during the session per area
  sum(stan_data$y[stan_data$area == a, k] > 0)
})
print(n_observed)
# model estimate
n_stats_k <- fit$summary("N") %>%
  filter(grepl(paste0(",", k, "]"), variable)) %>%
  mutate(Area = factor(1:6), Observed = n_observed)
mean_N_sessione <- mean(n_stats_k$mean)
ggplot(n_stats_k, aes(x = factor(Area))) +
  # Incertezza del modello
  geom_errorbar(aes(ymin = q5, ymax = q95), width = 0.2, color = "royalblue") +
  # Stima media del modello
  geom_point(aes(y = mean, color = "Abundance estimate (N)"), size = 4) +
  # Dato grezzo (Individui diversi osservati)
  geom_point(aes(y = Observed, color = "Raw data (observed)"), shape = 17, size = 4) +
  geom_hline(yintercept = mean_N_sessione, linetype = "dashed",
    color =  "darkgrey", size = 0.8) +
  annotate("text", x = 0.8, y = mean_N_sessione + 1,
    label = "Mean abundance", color = "darkgrey", hjust = 0) +
  scale_color_manual(values = c("Abundance estimate (N)" = "royalblue", "Raw data (observed)" = "red")) +
  labs(title = "Comparison between estimated and observed individuals",
       subtitle = "The difference represents the individuals estimated but not observed",
       x = "Study area", y = "Number of individuals") +
  theme_minimal()

```


```{r}
#| label: shrinkage-N-area
#| fig-cap: "Shrinkage: Naive vs Model"

# Estrai i dati grezzi dal tuo oggetto 'y' (la matrice di cattura)
# n_osservati deve essere un vettore di 23 elementi
n_osservati_area1 <- apply(stan_data$y[stan_data$area_idx == 1, ], 2, sum) 
n_osservati_area2 <- apply(stan_data$y[stan_data$area_idx == 2, ], 2, sum) 
n_osservati_area3 <- apply(stan_data$y[stan_data$area_idx == 3, ], 2, sum) 
n_osservati_area4 <- apply(stan_data$y[stan_data$area_idx == 4, ], 2, sum) 
n_osservati_area5 <- apply(stan_data$y[stan_data$area_idx == 5, ], 2, sum) 
n_osservati_area6 <- apply(stan_data$y[stan_data$area_idx == 6, ], 2, sum) 

# Estrai le stime del modello
n_stats <- vector("list", stan_data$N_areas)
n_stats[[1]] <- fit$summary("N") %>% 
  filter(grepl("N\\[1,", variable)) %>% 
  mutate(Sessione = 1:23, 
         Osservati = n_osservati_area1,
         Data = master_dates) # Le date che abbiamo definito prima
n_stats[[2]] <- fit$summary("N") %>% 
  filter(grepl("N\\[2,", variable)) %>% 
  mutate(Sessione = 1:23, 
         Osservati = n_osservati_area2,
         Data = master_dates) # Le date che abbiamo definito prima
n_stats[[3]] <- fit$summary("N") %>% 
  filter(grepl("N\\[3,", variable)) %>% 
  mutate(Sessione = 1:23, 
         Osservati = n_osservati_area3,
         Data = master_dates) # Le date che abbiamo definito prima
n_stats[[4]] <- fit$summary("N") %>% 
  filter(grepl("N\\[4,", variable)) %>% 
  mutate(Sessione = 1:23, 
         Osservati = n_osservati_area4,
         Data = master_dates) # Le date che abbiamo definito prima
n_stats[[5]] <- fit$summary("N") %>% 
  filter(grepl("N\\[5,", variable)) %>% 
  mutate(Sessione = 1:23, 
         Osservati = n_osservati_area5,
         Data = master_dates) # Le date che abbiamo definito prima
n_stats[[6]] <- fit$summary("N") %>% 
  filter(grepl("N\\[6,", variable)) %>% 
  mutate(Sessione = 1:23, 
         Osservati = n_osservati_area6,
         Data = master_dates) # Le date che abbiamo definito prima

plot_shr <- vector("list", stan_data$N_areas)
for (i in 1:stan_data$N_areas) {
  ggplot(n_stats[[i]], aes(x = Data)) +
    # Area di incertezza del modello
    geom_ribbon(aes(ymin = q5, ymax = q95), fill = "royalblue", alpha = 0.2) +
    # Linea della stima media N
    geom_line(aes(y = mean, color = "Stima Modello (N)"), size = 1) +
    # Punti dei dati grezzi osservati
    geom_point(aes(y = Osservati, color = "Catturati (Dato Grezzo)"), size = 2) +
    scale_color_manual(values = c("Stima Modello (N)" = "royalblue",
      "Catturati (Dato Grezzo)" = "red")) +
    labs(title = paste0("Shrinkage e Correzione dell'Abbondanza - Area", i),
       subtitle = "Confronto tra individui catturati e popolazione stimata per sessione",
       x = "Data Sessione", y = "Numero di Individui") +
    theme_minimal() -> plot_shr[[i]]
}
print(plot_shr[[1]])
print(plot_shr[[2]])
print(plot_shr[[3]])
print(plot_shr[[4]])
print(plot_shr[[5]])
print(plot_shr[[6]])
```

## 4. Prior sentivity checks

```{r}
#| label: prior-sensitivity-models

###---- Prior sensitivity ----
fit_mod_base <- readRDS("outputs/rds/fit_mod_base.rds")
fit_mod_inform <- readRDS("outputs/rds/fit_mod_inform.rds")
fit_mod_regular <- readRDS("outputs/rds/fit_mod_regular.rds")
fit_mod_strong <- readRDS("outputs/rds/fit_mod_strong.rds")
fit_mod_wide <- readRDS("outputs/rds/fit_mod_wide.rds")
```


```{r}
#| label: prior-sensitivity-comparison
#| fig-cap: "Prior sensitivity checks: Comparison of global parameters"

# Lista dei modelli per facilitare il loop
modelli <- list(
  "Base" = fit_mod_base,
  "Informative" = fit_mod_inform,
  "Regular" = fit_mod_regular,
  "Strong" = fit_mod_strong,
  "Wide" = fit_mod_wide
)

# Estrazione e unione dei parametri
df_forest <- lapply(names(modelli), function(nome) {
  modelli[[nome]]$summary(c("mu_active", "mu_hibernation", "alpha_p", "beta")) %>%
    mutate(Prior = nome)
}) %>% bind_rows()

# Pulizia nomi parametri per una migliore lettura
df_forest <- df_forest %>%
  mutate(variable = gsub("\\[.*\\]", "", variable)) # Rimuove gli indici [1,1] per raggruppare i tipi di parametro

# Pulizia e raggruppamento dei dati per il Forest Plot
df_forest_clean <- df_forest %>%
  # Estraiamo il nome base del parametro (es. da alpha_p[1,1] a alpha_p)
  mutate(param_type = gsub("\\[.*\\]", "", variable)) %>%
  # Raggruppiamo per modello e tipo di parametro
  group_by(Prior, param_type) %>%
  # Calcoliamo la media delle medie e degli intervalli per avere un punto unico
  summarise(
    mean_global = mean(mean),
    q5_global = mean(q5),
    q95_global = mean(q95),
    .groups = "drop"
  )

# Nuovo Plot pulito
ggplot(df_forest_clean, aes(x = mean_global, y = Prior, color = Prior)) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = q5_global, xmax = q95_global), height = 0.3, size = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~param_type, scales = "free_x") +
  labs(
    title = "Prior Sensitivity Check (Global Trends)",
    subtitle = "Aggregated posterior means across different prior specifications",
    x = "Posterior Mean (Logit scale for alpha_p, mortality scale for mu)",
    y = NULL
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none", strip.text = element_text(face = "bold", size = 12))

ggsave(
    filename = "figs/Prior_Sensitivity_checks.tiff", # o .png, .pdf, .eps
    plot = last_plot(),                          # esporta l'ultimo grafico visualizzato
    device = "tiff",                             # formato file
    compression = "lzw",
    width = 180,                                 # larghezza (es. 180mm è lo standard "full page width")
    height = 120,                                # altezza in mm
    units = "mm",                                # unità di misura
    dpi = 300,                                   # risoluzione richiesta
    bg = "white"                                 # sfondo bianco (evita trasparenze indesiderate)
  )
```

Mortality Rates ($\mu_{active}$ and $\mu_{hibernation}$): Posterior means for both active-season and winter mortality remain remarkably stable across all five models, from "Strong" to "Wide" priors. This demonstrates that the estimated mortality signal is robust. The data consistently identifies a monthly mortality rate around 0.25–0.30 for the active season and a lower instantaneous monthly rate for winter that, due to its 7-month duration, results in significant cumulative loss.

Detection Probability ($\alpha_p$): While the "Wide" model shows a much larger uncertainty (wider credible intervals), the central estimate remains consistent with the more informative models. This justifies the use of weakly informative priors in the base model. The data alone struggle to "anchor" detectability precisely, but the priors help stabilize the model without biasing the final demographic trend.

Recruitment/Entrance ($\beta$): Estimates for recruitment remain consistent regardless of the prior's strength. The model reliably identifies the rate at which new individuals enter the population, ensuring that the observed decline is a true reflection of the balance between survival and recruitment.

Overall, the overlap of the posterior distributions across varying levels of prior informativeness proves that the model is well-identified.

#### Overlap of posteriors

```{r}
#| label: prior-sensitivity-overlaps
#| fig-cap: "Prior sensitivity checks: Comparison of global parameters"


# 1. Funzione universale per confrontare i 5 modelli su un singolo parametro
compare_5_priors_flexible <- function(param_base, area_idx = NULL, session_idx = NULL, title_label) {
  
  # Costruisci il nome del parametro: se gli indici sono forniti li usa, altrimenti no
  if (!is.null(area_idx) && !is.null(session_idx)) {
    vble <- paste0(param_base, "[", area_idx, ",", session_idx, "]")
  } else if (!is.null(area_idx)) {
    vble <- paste0(param_base, "[", area_idx, "]")
  } else {
    vble <- param_base
  }
  
  # Estrazione dei draws dai 5 modelli (usa i tuoi oggetti fit_mod_...)
  draws_list <- list(
    "Base" = as.numeric(fit_mod_base$draws(vble)),
    "Informative" = as.numeric(fit_mod_inform$draws(vble)),
    "Regular" = as.numeric(fit_mod_regular$draws(vble)),
    "Strong" = as.numeric(fit_mod_strong$draws(vble)),
    "Wide" = as.numeric(fit_mod_wide$draws(vble))
  )
  
  # Calcolo Overlap Numerico
  cat(sprintf("\n--- Overlap Analysis: %s ---\n", vble))
  for(n in names(draws_list)[-1]) {
    ov <- round(overlapping::overlap(list(draws_list[["Base"]], draws_list[[n]]))$OV, 2)
    cat(sprintf("Base vs %s: %s\n", n, ov))
  }
  
  # Plot
  df_plot <- data.frame(
    value = unlist(draws_list),
    Prior = rep(names(draws_list), each = length(draws_list[[1]]))
  )
  
  ggplot(df_plot, aes(x = value, fill = Prior)) +
    geom_density(alpha = 0.3) +
    labs(title = title_label, subtitle = paste("Parameter:", vble),
         x = "Estimated value (Posterior)", y = "Density") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set1")
}

# Reclutamento Area 1, Sessione 12
plot_beta_check <- compare_5_priors_flexible()("beta", 1, 12, "Prior Sensitivity: Recruitment (Beta)")
plot_beta_check
plot_beta_check <- compare_5_priors_area_session("beta", 6, 12, "Prior Sensitivity: Recruitment (Beta)")
plot_beta_check

plot_alpha_p_check<- compare_5_priors_area_session("alpha_p", 1, 12, "Prior Sensitivity: Detectability (p)")
plot_alpha_p_check


plot_mu_hibernation <- compare_5_priors_flexible("mu_hibernation", title_label = "Global Mortality Overlap - Hibernation period")
plot_mu_hibernation
plot_mu_active <- compare_5_priors_flexible("mu_active", title_label = "Global Mortality Overlap - Active period")
plot_mu_active
```


